\documentclass{article}


\title{Thoughts on Hooks in code}
\author{Frank Mittelbach \and
%                        \and  
%                        \and  
  \LaTeX{} Project Team}

\date{2019-10-15}

\newcounter{hook}
\newcounter{config}

\newenvironment{hook}[3][new]  % new/exists, area, name 
   {\refstepcounter{hook}%
    \begin{itemize}%
    \item[\textbf{Hook (\thehook):}]\textbf{#3} (#2/\textit{#1})}
   {\end{itemize}}

\newenvironment{config}[3][new]  % new/exists, area, name 
   {\refstepcounter{config}%
    \begin{itemize}%
    \item[\textbf{Config (\theconfig):}]\textbf{#3} (#2/\textit{#1})}
   {\end{itemize}}

\newenvironment{comment}[2][]  % initial / date
               {\par[#1 #2]\itshape}
               {\par}
   
\newcommand\cs[1]{\texttt{\textbackslash #1}}               
\newcommand\pkg[1]{\texttt{#1}}

\renewcommand\arg[1]{\textit{#1}}      % don't need the math Op in this document I hope :-)

\usepackage[T1]{fontenc}
\usepackage{csquotes}

\begin{document}

\maketitle

\tableofcontents

\section{Introduction}


This document collects some thoughts on places for hooks (i.e., code
interfaces where other code can add material for execution).

I distinguish this from what I call  ``configuration points'', which I think of as
being commands that have a single definition that can be replaced by a
different one but not ``added to'' by different packages. For example,
``add begin document'' is a typical hook, as it is a place where
different packages my want to execute some code. On the other hand
``prepare footins'' is a configuration interface whose sole purpose is
to do something specific to the \cs{footins} box. This may differ
based on the intended outcome, but there can only one definition active at any
time.\footnote{Technically this is of course not quite true, there is
  a certain gray area, but only if some packages very closely
  interrelated. The main point is that ``hooks'' are more about places
  where code gets added without (much) concern about what other
  package want to execute.}


\subsection{Hooks and their interface}

At the moment most existing hooks in \LaTeXe{} are actually named
\cs{@...hook} (with one or two exception). We should keep that
approach for backward compatibility and only change the exceptions.

Currently we do have declarations such as \cs{AtBeginDocument} to add
data to a hook.  The order in which the code gets executed then only
depends on the order of the declarations, i.e., on the order of
package loading.

What is not possible right now is define known dependencies between
different code executed in a certain place, result in the famous ``try
changing the order of package X and Y to resolve the
problem''. Sometimes this works, but sometime a different order would
be needed for different hooks and then of course that advice doesn't
really any longer work.

As an alternative/extension we envision that material added to a hook
will be named (using the package name by default) and that it will be
possible for packages to declare rules that its code for a certain
hook should be placed in relation to code from some other package (if
that package is loaded). This way known dependencies could be
automatically resolved (if declared as a rule by either side---or even
both if they don't conflict) and impossible requirements could be
detected as being incompatible.

Outline of interface:
\begin{verbatim}
 \AddToHook{<hook>}{<name>}{<code>}
 \ClearHook{<hook>}{<name>}              % maybe
 \SetupHook{<hook>}{<name>}{<code>}      % \ClearHook followed by \AddToHook

 \SetupHookRule{<hook>}{<name1>}{<name2>}{<rule>}
\end{verbatim}

\cs{AddToHook} would add additional \arg{code} labeled by \arg{name}
to the \arg{hook}. If there exists already code under that \arg{name}
then it is appended.  \cs{ClearHook} removes all code labeled by
\arg{name}.  \cs{SetupHookRule} defines some relationship between
\arg{name1} and \arg{name2}, for the exact syntax a lot different ways
are possible (should be determined later).

One could also think of \arg{name} to be optional defaulting to the
current package/class name to encourage people to use that as the name
normally.

[2019/06/30] One important aspect not covered above is the question of
what is stored with respect to \arg{code}. In most cases it should be
the unexpanded code, but we will also need a version of all commands
above that stored the expansion (protected expansion) in case values
current at the time of setup need to be stored.


Existing hook interface commands such as \cs{AtBeginDocument} would
still be supported as follows:
\begin{itemize}
\item
  They add their code under the name ``\texttt{legacy}'' to the hook.
\item
  Other packages could then decide that they should come before or
  after the code labeled \texttt{legacy}.
\end{itemize}
In other words:
\begin{verbatim}
 \def\AtBeginDocument#1{\AddToHook{begindocument}{legacy}{#1}}
 \def\AtEnddocument#1{\AddToHook{enddocument}{legacy}{#1}}
\end{verbatim}

However, there use would be discouraged in favor of explicitly
labeling the hook code with the package name.

There also exist \cs{AtEndOfPackage} and \cs{AtEndOfClass} which uses
\begin{verbatim}
 \@currname.\@currext-h@@k
\end{verbatim}
as the hook name, but I don't think
this needs to be incorporated into the more general mechanism as I
don't think this being usable anywhere outside the actual package
code---that one package writes into that hook of another package is in
theory be possible (as in overwrite some code of that package if it
gets loaded later than me, but I doubt that this makes a
useful/explainable interface).  On the other hand it is surprisingly
often used (in 379 package/classes in TL 2019) so perhaps that
assumption needs some checking.


\subsection{Configuration points and their interface}

As configuration points allow for only a single definition at any one
time they can be simply defined as commands without providing an
explicit interface.

My current suggestion is to all call them \cs{@...config} in the
\LaTeXe{} code and provide them with a default definition. We could
think of offering a configuration interface such as
\begin{verbatim}
 \SetupConfigPoint{@makecol}{preparefootins}{<code>}
\end{verbatim}
but that wouldn't do much other than (globally) replacing the config
command definition with the new one.

I also think that such config point commands should not take arguments
even if they technically have inputs (like the current page box or the
\cs{footins} box etc.).



\subsection{Distinction between configuration points and hooks}

As a rule of thumb, any place where the result of executing \arg{code}
requires a certain specific outcome then it is normally a
configuration point and not a general hook, e.g., \enquote{the code
  write something into a certain box} or \enquote{the code runs a
  specific set of other code fragments and all it is supposed to do is
  to decide the order or not run some of them}.  There could be
excepts and boundary cases but I think as a basic rule this makes
sense.

In contrast if we don't make any (or not much) assumptions on what
could be executed and what the side effects are supposed to be then it
is most likely a general hook.




\subsection{Thoughts on current \texttt{l3hooks} interfaces}

In \texttt{l3hooks} we currently have the following type of interfaces
that would need some merging with the thoughts above.

\subsubsection{Next viz all (or what is called \enquote{repeated hooks})}

First of all there is the idea of executing some code only once a hook
is reached the next time, e.g., \enquote{do something on the next
  page} in contrast to \enquote{do something on every page}. The
question to resolve here seems to me what is the relationship between
code of the different types. Probably the easiest way to handle this
is to execute \enquote{next} code always after \enquote{all} code.

However, is this general enough or is more control needed?


\subsubsection{Document viz repeated hooks}

In short I'm not yet convinced that the distinction is useful, is it?
If so how/why? Example please.

\begin{comment}[JAW]{2019-07-01}
  The starting point for the split is that it makes sense to namespace
  document-wide hooks in one place: it does not really matter which
  module adds them. That of course could be handled by
\begin{verbatim}
    \hook_new:nn { document } { name-of-hook }
\end{verbatim}

The second thing is that the next/all mechanism is redundant for such
hooks. So I thought I'd just make a dedicated interface that did not
have quite the same overhead.

The third and perhaps most important thing is that document hooks can
only be used once and this affects adding to them. If you look at
current \cs{AtBeginDocument}, it can still be 'added to' after \cs{document},
but is a simple \cs{@firstofone}. To do that, the hook needs to 'know'
it's a one-shot. I was aiming for something similar, although I've
currently gone for an error rather than silently dumping the
tokens. My thinking is that
\begin{verbatim}
    \hook_use:nn { document } { name }
    \hook_gpush:nnnn { document } { name } { pkg } { content }
\end{verbatim}
is going to be quite surprising.
\end{comment}



\subsection{Thoughts on the \cs{end} (environment) interface}

From a mail by Will:
\begin{verbatim}
1.  It’s interesting that breqn redefines \end or now “\end “ to be:

      \@namedef{end }#1{\csname end#1\endcsname \latex@end{#1}}%

This allows an environment end definition like this:

      \def\enddmath#1{\check@punct@or@qed}

which “absorbs” the \latex@end (the #1), switches over to the aux
 function, which is

      \def\check@punct@or@qed#1{…
         \def\finish@end{\csname end@#1\endcsname\latex@end{#1}}%
         \check@punct
      }

which allows \futurelet lookahead and all that, using \finish@end
 finally to tidy up the environment properly.

This is all rather convoluted and very neat and tidy.

As far as I know there are no incompatibilities with breqn and any
 other major package, so the question arises whether it would be worth
 setting up "\end “ a little more directly to allow this behaviour. Of
 course it will slow down every environment in a document by one
 expansion, but is that a big deal? It may well be, I don’t know.


2.  If we build a hook into 2e to provide this, do we add an interface
 to xparse to allow environments to lookahead? As well as being useful
 for breqn — and it’s such a good part of breqn I’d argue it’s worth
 breaking it out from there for xmath — wouldn’t this style of
 lookahead be needed for LDB concepts?
\end{verbatim}


\section{Hooks and config points in various places}

In this section we collect existing \LaTeXe{} hooks (both by the
kernel and/or by package) as well a hooks that do not exist but would
be beneficial to have for one or the other reason. As of now it mainly
discusses the \LaTeXe{} situation.

For each hook/config point we document
\begin{itemize}
\item a ``name'';
\item the area where it applies;
\item whether it is ``new'' or does already exist (if so where);
\item and the major reason why it would be beneficial.
\end{itemize}
If there are several distinct use cases each could be added using
\cs{item}. You can cross reference hooks using
\cs{label}/\cs{ref}.



\subsection{Document structure}



\begin{hook}[kernel]{doc-structure}{documentclass}
\item
  This is really more or less internal (used for handling 2.09
  compatibility) but it shows up in one or two other places in TL.
\end{hook}

\begin{hook}[kernel]{\cs{document}}{begindocument}
\item
\end{hook}

\begin{hook}[kernel]{\cs{enddocument}}{enddocument}
\item
\end{hook}



\subsection{Output routine}

\subsubsection{Making pages}


\begin{hook}[kernel]{\cs{@outputpage}}{begindvibox}
\item
  \cs{AtBeginDvi} is a legacy \LaTeXe{} interface that hooks into the
  first shipout box at the very top. Afterwards the material is
  dropped so it doesn't show up in later boxes. What is special is
  that the hook itself is implemented as a box and each time
  \cs{AtBeginDvi} is called more material is added to that box which
  eventually is unboxed into the first shipout box. in that respect
  that hook is rather ``special'' and right now I'm not sure it really
  has to---use cases please.
\item
  If not then my proposal would be to deprecate it in favor of a
  ``normal'' hook at the same place (i.e., implemented as a token
  list) that would fit into the general model outlined above.
\item
  [2019/06/30] The current thinking is that the use of a box for
  storage was mainly done to help preserving space in the early
  \LaTeXe{} days, so it should be possible to drop that in favor of a
  hook where the code is stored like any other hook in form of a token
  list.
\end{hook}

\begin{hook}{\cs{@outputpage}}{firstshipout}
\item
  Suggested replacement for \texttt{begindvibox} as a normal named
  hook. Executed at the very top inside the first shipoutbox.
\item
  Not being implemented as a box register internally also means that
  it could execute other code than just adding \cs{special}s etc to
  the shipout box.
\end{hook}

\begin{hook}{\cs{@outputpage}}{shipout}
\item
  Executed directly at the top of the shiout box but only for shipouts
  after the first (which uses \texttt{firstshipout}.
\item
  For code that should be executed at all shipouts one need to put it
  into both \texttt{firstshipout} and \texttt{shipout}.
\end{hook}

The above is probably not general enough in the light of what
\pkg{atbegshi} implements (fill in correct details \ldots{} not done):

\begin{hook}[atbegshi]{\cs{@outputpage}}{shipoutbox}
\item
\end{hook}

\begin{hook}[atbegshi]{\cs{@outputpage}}{forgroundshipoutbox}
\item
\end{hook}



\subsubsection{Making columns}

\begin{hook}{\cs{@makecol}}{pre}
\item
  \pkg{manyfoot} and similar packages like to take control at the very
  beginning of column generation.
\end{hook}

\begin{hook}{\cs{@makecol}}{post}
\item
  And we may have some post-processing going on after the column is
  made up.
\end{hook}

\begin{config}{\cs{@makecol}}{footins}
\item
  To support manipulation of footnote text (like footnotes as paras,
  in 2 columns etc).

  At this point the assembled footnotes are inside box \cs{footins} and any
  manipulation needs to globally write them back in there.
\end{config}

\begin{config}{\cs{@makecol}}{blocks}
\item
  A column supports the following block elements:
  \begin{itemize}
  \item
    galley text (already inside \cs{@outputbox})
  \item
    footnotes (initially inside \cs{footins})
  \item
    top and bottom floats (initially inside the \LaTeXe{} float lists
    for top and bottom)
  \end{itemize}
  These can appear in different order within the column and in this
  config point the ordering and any special spacing between them is
  specified. The final result is stored in \cs{@outputbox}.

  For specification only the following commands should be used:
  \begin{itemize}
  \item \cs{@makecolappendfootnotes}\arg{separation} The separation
    has to be \cs{vfill} or \cs{@makecol@moveskip} (which is not
    quite the right interface).
  \item \cs{@makecolattachfloats}
  \item \cs{@makecolappendstuff}\arg{vertical material}
  \end{itemize}
  For example to get the footnotes above the bottom floats one would specify:
\begin{verbatim}
 \SetupConfigPoint{@makecol}{blocks}
                  {\@makecol@appendfootnotes {\vfill}%
                   \@makecol@appendfloats}
\end{verbatim}


  Discussion notes:
  \begin{itemize}
  \item
    Maybe \@outputbox{} should start out empty there should be an
    explicit \cs{makecolappendtext}
  \item
    Technically attaching top and bottom floats in one go should be
    enough but perhaps it would be clearer if there are two commands
    \cs{@makecolappendtopfloats} and \cs{@makecolappendbottomfloats}
    even though their order should always be top above bottom.
  \item
    There have to be some clear rules on how \cs{@makecolappendstuff}
    can be used.
  \item
    Should spaces be automatically suppressed in configuration point setups?
  \end{itemize}
\end{config}

\begin{config}{\cs{@makecol}}{splitfootnotemessage}
\item
  Executes if a footnote gets split. By default empty but could be
  used to set up a warning message or similar in that case.
\end{config}


\subsection{Heading commands}

\begin{hook}{heading}{before-break}
\item
  For heading commands it is helpful if one can issue \cs{mark}
  commands, e.g. \cs{PutMark} from \pkg{xmarks} to support running
  header or footer setup. This has to be possible directly after
  heading title (where it is currently supported through commands like
  \cs{sectionmark}) but also directly before the break to determine
  the placement of the heading (e.g., at the very top of a page/column
  or elsewhere). The mark needs to come after the heading has done its
  magic with any preceding space, it can't hide such space from the
  command. In other words it need to happen basically in the middle of
  \cs{addpenalty}.
\item
  For L3 the galley concept might be able to handle this properly, but
  most likely also need more than one point of code injection.
\end{hook}



%\subsection{}

\end{document}
