% Copyright 2008 The LaTeX Project
\documentclass{ltnews}
\renewcommand{\LaTeXNews}{\LaTeX3~News}

\publicationmonth{January}
\publicationyear{2009}
\publicationissue{1}

\begin{document}
\maketitle

\section{Welcome to \LaTeX3}

Momentum is slowly starting to build again behind the \LaTeX3 project. For the
last few releases of \TeX~Live, the experimental programming foundation for
\LaTeX3 has been available under the name \package{expl3}. Despite large
warnings that the code would probably change in the future, we wanted to show
that there was progress being made, no matter how slowly. Since then, some
people have looked at the code, provided feedback, and~--- most
importantly~--- actually tried using it. Although it is yet early days, we
believe that the ideas behind the code are sound and there are only `cosmetic
improvements' that need to be made before \package{expl3} is ready for the
\LaTeX~package author masses.

\section{What currently exists}

The current state of the \LaTeX3 code consists of two main branches: the
\package{expl3} modules that define the underlying programming environment,
and the `\package{xpackage}s', which are a suite of packages that are written
with the \package{expl3} programming interface and provide some higher-level
functionality for what will one day become \LaTeX3 proper. \package{expl3} and
parts of the \package{xpackages} are designed to be used \emph{on top} of
\LaTeXe, so new packages can take advantage of the new features while still
allowing to be used many of the vast number of \LaTeXe\ packages on CTAN.

\section{What's happening now}

In preparation for a refactor of the \package{expl3} code, a comprehensive
test suite is being written in order to provide the safety of regression tests
for the code that we end up changing. As these tests are being written, our
assumptions about what should be called what and the underlying naming
conventions for the functions and datatypes defined in \LaTeX3 are being
questioned, challenged, and noted for further rumination.

At time of writing, we are approximately half-way through writing the test
suite. Once this task is complete, which we hope to take place in the first
half of 2009, we will be ready to make changes without worrying about breaking
anything.

\section{What's happening soon}

So what do we want to change? The current \package{expl3} codebase has
portions that literally date to the pre-\LaTeXe\ days, while other modules
have been more recently conceived. It is quite apparent when reading through
the sources that some unification and tidying up would improve the simplicity
and consistency of the code. In many cases, such changes will mean nothing
more than a tweak or a rename.

Beyond these minor changes, we are also re-thinking the exact notation behind
the way functions are defined. There are currently a handful of different
types of arguments that functions may be passed (from an untouched single
token to a complete expansion of a token list) and we're not entirely happy
with how the original choices have evolved now that the system has grown
somewhat. We have received good feedback from several people on ways that we
could improve the argument syntax, and as part of the refactoring of the
\package{expl3} packages we hope to address the problems that we currently
perceive in the present syntax.

\section{What's happening later}

It's a little early to start talking about what will happen when the
refactoring process has finished. We hope to start freezing the interface at
that point, and we hope that more package authors will be interested in using
the new ideas to write their own code. The scope of features provided by the
\package{expl3} modules will only continue to grow as it is used more and more
in the real world.

Some new and/or experimental packages will be moving to use the
\package{expl3} programming interface, including \package{breqn},
\package{mathtools}, \package{empheq}, \package{fontspec}, and
\package{unicode-math}. (Which is one reason towards the lack of progress in
these latter two in recent times.) There will also be a version of the
\package{siunitx} package written in \package{expl3}, in parallel to the
current \LaTeXe\ version. These developments will provide improvements to
everyday \LaTeX\ users who haven't even heard of the \LaTeX3 Project.

Looking towards the long term, \LaTeX3 as a document preparation system needs
to be written almost from scratch. A high-level user syntax needs to be
designed and scores
of packages will be incorporated into the `out-of-the-box' default document
templates. \LaTeXe\ has stood up to the test of time~--- some fifteen years
and still going strong~--- and it is now time to write a successor that will
survive another score.

\end{document}
 
















