\iffalse meta-comment

File: l3styleguide.tex

Copyright (C) 2011,2012,2014-2025 The LaTeX Project

It may be distributed and/or modified under the conditions of the
LaTeX Project Public License (LPPL), either version 1.3c of this
license or (at your option) any later version.  The latest version
of this license is in the file

   https://www.latex-project.org/lppl.txt

This file is part of the "l3kernel bundle" (The Work in LPPL)
and all files in that bundle must be distributed together.

The released version of this bundle is available from CTAN.

\fi

\documentclass{l3doc}


\title{%
  The \LaTeX3 kernel: style guide for code authors%
}
\author{%
  The \LaTeX{} Project\thanks
    {%
      E-mail:
      \href{mailto:latex-team@latex-project.org}%
        {latex-team@latex-project.org}%
    }%
}
\date{Released 2025-06-09}

\begin{document}

\maketitle

\tableofcontents

\section{Introduction}

This document is intended as a style guide for authors of code and
documentation for the \LaTeX3 kernel. It covers both aspects of coding
style and the formatting of the sources. The aim of providing these
guidelines is help ensure consistency of the code and sources from
different authors. Experience suggests that in the long-term this helps
with maintenance. There will of course be places where there are
exceptions to these guidelines: common sense should always be
applied!

\section{Documentation style}

\LaTeX3 source and documentation should be written using the document
class \cls{l3doc} in \file{dtx} format. This class provides a number
of logical mark up elements, which should be used where possible.
In the main, this is standard \LaTeX{} practice, but there are a
few points to highlight:
\begin{itemize}
  \item
    Where possible, use \cs{cs} to mark up control sequences
    rather than using a verbatim environment.
  \item
    Arguments which are given in braces should be marked using
    \cs{Arg} when code-level functions are discussed, but using
    \cs{marg} for document functions.
  \item
    The names \TeX{}, \LaTeX{}, \emph{etc}.\ use the normal logical mark
    up followed by an empty group (|{}|), with the exception of |\LaTeX3|,
    where the number should follow directly.
  \item
    Where in line verbatim text is used, it should be marked up
    using the \verb=|...|= construct (i.e., vertical bars delimit
    the verbatim text).
  \item In line quotes should be marked up using the \cs{enquote}
    function.
  \item
    Where numbers in the source have a mathematical meaning,
    they should be included in math mode. Such in-line math mode
    material should be marked up using |$...$| and  \emph{not}
    |\(...\)|.
\end{itemize}

Line length in the source files should be under $80$
characters where possible, as this helps keep everything on the screen
when editing files. In the \file{dtx} format, documentation lines start
with a \texttt{\%}, which is usually followed by a space to leave a
\enquote{comment margin} at the start of each line.

As with code indenting (see later), nested environments and arguments
should be indented by (at least) two spaces to make the nature of the nesting
clear. Thus for example a typical arrangement for the \env{function}
environment might be
\begin{verbatim*}
\begin{function}{\seq_gclear:N, \seq_gclear:c}
  \begin{syntax}
    \cs{seq_gclear:N} \meta{sequence}
  \end{syntax}
  Clears all entries from the \meta{sequence} globally.
\end{function}
\end{verbatim*}
The \enquote{outer} \verb*|% \begin{function}| should have the customary
space after the |%| character at the start of the line.

In general, a single \env{function}  or \env{macro} environment should be
used for a group of closely-related functions, for example argument
specification variants. In such cases, a comma-separated list should be
used, as shown in the preceding example.

\section{Format of the code itself}

The requirement for fewer than $80$ characters per line applies to the code
itself as well as the surrounding documentation. A number of the general
style principles for \LaTeX3 code apply: these are described in the following
paragraph and an example is then given.

With the exception of simple runs of parameter (|{#1}|, |#1#2|,
etc.), everything should be divided up using spaces to make the code
more readable. In general, these will be single spaces, but in some
places it makes more sense to align parts of the code to emphasize
similarity. (Tabs should not be used for introducing white space.)

Each conceptually-separate step in a function should be on a separate
line, to make the meaning clearer. Hence the \texttt{false} branch
in the example uses two lines for the two auxiliary function uses.

Within the definition, a two-space indent should be used to show each
\enquote{level} of code. Thus in the example \cs{tl_if_empty:nTF} is
indented by two spaces, but the two branches are indented by four
spaces. Within the \texttt{false} branch, the need for multiple lines
means that an additional two-space indent should be used to show that
these lines are all part of the brace group.

The result of these lay-out conventions is code which in general
looks like the example:
\begin{verbatim*}
\cs_new:Npn \module_foo:nn #1#2
  {
    \tl_if_empty:nTF {#1}
      { \module_foo_aux:n { X #2 } }
      {
        \module_foo_aux:nn {#1} {#2}
        \module_foo_aux:n { #1 #2 }
      }
  }
\end{verbatim*}

\section{Code conventions}

All code-level functions should be \enquote{long} if they accept any
arguments, even if it seems \enquote{very unlikely} that a \cs{par} token
will be passed. Thus \cs{cs_new_nopar:Npn} and so forth should only be used
to create interfaces at the document level (where trapping \cs{par} tokens
may be appropriate) or where comparison to other code known not to be
\enquote{long} is required (\emph{e.g.}~when working with mixed
\LaTeXe{}/\pkg{expl3} situations).

The expandability of each function should be well-defined. Functions which
cannot be fully expanded must be \texttt{protected}. This means that expandable
functions must themselves only contain expandable material. Functions which
use any non-expandable material must be defined using \cs{cs_new_protected:Npn}
or similar.

When using \cs{cs_generate_variant:Nn}, group related variants together
to make the pattern clearer. A common example is variants of a function
which has an \texttt{N}-type first argument:
\begin{verbatim}
  \cs_generate_variant:Nn \foo:Nn {     NV , No }
  \cs_generate_variant:Nn \foo:Nn { c , cV , co }
\end{verbatim}

There may be cases where omitting braces from \texttt{o}-type arguments
is desirable for performance reasons. This should only be done if the
argument is a single token, thus for example
\begin{verbatim}
  \tl_set:No \l_some_tl \l_some_other_tl
\end{verbatim}
remains clear and can be used where appropriate.

\section{Private and internal functions}

Private functions (those starting \cs{__}) should not be used between modules.
The only exception is where a \enquote{family} of modules share some
\enquote{internal} methods: this happens most obviously in the kernel itself.
Any internal functions or variables \emph{must} be documented in the same way
as public ones.

The \pkg{DocStrip} method should be used for internal functions in a module.
This requires a line
\begin{quote}
  \ttfamily
  \%<@@=\meta{module}>
\end{quote}
at the start of the source (\texttt{.dtx}) file, with internal functions
then written in the form
\begin{verbatim}
  \cs_new_protected:Npn \@@_function:nn #1#2
    ...
\end{verbatim}

\subsection{Access from other modules}

There may be cases where it is useful to use an internal function from
a third-party module (this includes cases where you are the author of both
but they are not part of the same \enquote{family}). In these cases, you should
\emph{copy} the definition of the internal function to your code: this avoids
relying on non-documented interfaces. At the same time, it is strongly
encouraged that you discuss your requirements with the author of the
code you need to access. The best long-term solution to these cases is for
new documented interfaces to be added to the parent module.

\subsection{Access to primitives}

As \pkg{expl3} is still a developing system, there are places where direct
access to engine primitives is required. These are all marked as
\enquote{do not use} in the code and so require special handling. Where a
programmer is sure that they need to use a primitive (for example where the
team have not yet covered access to an area) then a local copy of the
primitive should be made, for example
\begin{verbatim}
  \cs_new_eq:NN \__module_message:w \tex_message:D
  % ...
  \cs_new_protected:Npn \__module_fancy_msg:n #1
    { \__module_message:w { *** #1 *** } }
\end{verbatim}
This approach makes it possible for the team and others to find such
usage (by searching for the \texttt{:D} argument type) but avoids
multiple uses in general code.

At the same time, the team ask that these use cases are raised by
emailing us. The team are keen to collect use cases for
areas that have not yet been addressed and to provide new code where the
required interfaces become clear.

Programmers using primitives should be ready to make updates to their
code as the team develop additional interfaces.

\section{Auxiliary functions}

In general, the team encourages the use of descriptive names in \LaTeX3 code.
Thus many helper functions would have names which describe briefly what they do,
rather than simply indicating that they are auxiliary to some higher-level
function. However, there are places where one or more \texttt{aux} functions
are required. Where possible, these should be differentiated by signature
\begin{verbatim}
  \cs_new_protected:Npn \@@_function:nn #1#2
    {
      ...
    }
  \cs_new_protected:Npn \@@_function_aux:nn #1#2
    {
      ...
    }
  \cs_new_protected:Npn \@@_function_aux:w #1#2 \q_stop
    {
      ...
    }
\end{verbatim}
Where more than one auxiliary shares the same signature, the recommended naming
scheme is \texttt{auxi}, \texttt{auxii} and so on.
\begin{verbatim}
  \cs_new_protected:Npn \@@_function_auxi:nn #1#2
    {
      ...
    }
  \cs_new_protected:Npn \@@_function_auxii:nn #1#2
    {
      ...
    }
\end{verbatim}
The use of \texttt{aux_i}, \texttt{aux_ii}, etc.\ is discouraged as this
conflicts with the convention used by \cs{use_i:nn} and related functions.

\section{Functions with `weird' arguments}

When defining commands that do not follow the usual convention of accepting
arguments as single-tokens or braced-text, the \verb|w| argument specifier
is used to denote that the function signature cannot fully describe the syntax.
Two examples from the \LaTeX3 kernel are:
\begin{quote}
\verb|\use_none_delimit_by_q_stop:w| $\langle$\,\emph{text}\,$\rangle$ \verb|\q_stop|\\
\verb|\use_i_delimit_by_q_stop:nw| \char`\{ $\langle$\,\emph{arg}\,$\rangle$\char`}\,$\langle$\,\emph{text}\,$\rangle$ \verb|\q_stop|
\end{quote}
More complex definitions are possible if commands are to parse tokens,
such as the internal kernel command
\begin{verbatim}
    \cs_new_protected:Npn \__clist_get:wN #1 , #2 \q_stop #3
      { \tl_set:Nn #3 {#1} }
\end{verbatim}
When the \verb|w| specifier is being used, it is encouraged not to try
and complicate the rest of the signature too much---for example, it
would be considered poor style to have a function with a signature like
\verb|\foo_bar:wnw| unless there were very clear reasons of code clarity.
A signature such as \verb|:ww| would certainly be discouraged.

Examining the examples above, it can be seen that there are scenarios in
which it may make logical sense for having a signature such as \verb|:wN| or
\verb|:nw|, but when in doubt the recommended approach is to simply use
\verb|:w| as a catch-all.

\end{document}
