%
% 
%
\typeout{ABSOLUTELY UNFINISHED!!!!!}
%

\documentclass{ltugboat}

\usepackage{calc}

% \verbfile{<name>} to input a possibly long file verbatim
%
\makeatletter
\def\verbfile#1{\begingroup
  \footnotesize
  \@verbatim
  \frenchspacing \@vobeyspaces
  \@input{#1}\endtrivlist\endgroup\@doendpe}
\makeatother

%\setcounter{secnumdepth}{-1}

\title{The \LaTeX{} regression test suite: concepts and
implementation}

\author{D. P. Carlisle \and  F. Mittelbach}

\begin{document}
\maketitle

\begin{abstract}
  This paper describes the concepts and the implementation of a
  regression test suite for \LaTeX{}. A general overview about the
  history behind this package is given in~\cite{tub:xxx}.

  As experience shows that there can't be enough test files in such a
  suite we make a plea to the \TeX{} community to help us making
  \LaTeX{} distributions even more reliable by joining a new volunteer
  group working on the task of updating and adding to this suite.
\end{abstract}

\section{Aims of this Package}
This set of files implements a scheme suggested by Frank Mittelbach
for ensuring that new releases of \LaTeX{} do not inadvertently modify
the behaviour of commands. This is definitely a non-trivial task, as
\LaTeX{} is a large system, and in `fixing' one bug, one often needs
to modify the definitions of several `internal' commands, which may
affect many other commands which have no obvious connection with the
original problem.

\section{The Basic System}
The idea is to have a suite of test files, each one exersising a
particular set of related commands. These should be called in such a
way that one can tell from the \texttt{.log} file (not the \texttt{.dvi}
file) that the command has `met its specification'. The \texttt{.log}
file is then edited to remove certain irrelevant information and will
then be stored, as a \texttt{.tlg} file. Before a new release is issued,
all the test files will be run through the new version, and the
resulting \texttt{.tlg} files will be automatically compared with the
saved original versions. Any tests which do not produce identical
results will then be notified to the maintainer of
\LaTeX, who can visually compare the \texttt{.tlg} files to see whether the
differences are due to an `improvement' or are the result of a newly
introduced bug! 

To distinguish test files from other files they have to have the
extension \texttt{.lvt} (standing for \LaTeX{} validation test). 


\section{The Format of a Test File}

If the functions to be tested need to be called inside the \LaTeX{}
\texttt{document} environment, then start the file as usual with:
\begin{verbatim}
\documentclass{article}
\begin{document}
\end{verbatim}
or any other class. In some cases you may also want to load some
packages if they are need for your test.

If you are testing some commands of the commands from \LaTeX`s kernel,
then these lines may be omitted.

Then start the test procedure with
\begin{verbatim}
\input{test2e}
\START
\end{verbatim}
Everything in the \texttt{.log} file before this will be omitted,
\verb|\START| also prints a character table in the \texttt{.log} file,
so that the \texttt{.tlg} file may be checked for errors caused be
faulty email connections.  After \verb|\START|, the \texttt{@} sign is
set up to act as a letter, so you can use any commands that can
normally only be used in a package or class file.

If you like you can identify yourself as the original author of this
test file by issuing the following commands.
\begin{verbatim}
\AUTHOR{my name}
\ADDRESS{my email address}
\end{verbatim}
This will help us to get back to you later if necessary.


You should then declare the format that you are using, the class, and
any packages that you have input. This should include dates and/or
version numbers, so that if at some future time, the test fails, any
differences between the original versions and the new versions can be
checked, to find the cause of the failure.

This is important as
not all packages declare themselves to the log file, and we can not
rely on TeX`s output as it includes full path names, and does not
include version numbers etc.  So for each package included give a
declaration like:
\begin{verbatim}
\FORMAT{LaTeX2e <1997/12/01> patch level 1}
\CLASS[a4paper]{article 1996/10/31 v1.3u}
\PACKAGE{ifthen 1996/08/02 v1.0m}
\PACKAGE[dvips]{graphics 1996/10/31 v1.0c}
\PACKAGES{array 1996/06/14 v2.3i}
\end{verbatim}

You may then optionally `declare' a collection of commands that make
up the `module' that you are testing.  \verb|\DECLARECOMMAND\foo| will
report whether \verb|\foo| is defined, undefined, or equivalent to
\verb|\relax|.  Note that you may also declare commands which are not
currently defined in \LaTeX, but which might be added. The system will
then pick up the addition of such a command, and remind the
maintainers to announce that there has been a change in the user
interface. Thus if you were testing the \texttt{array} environment,
you might want to declare \verb|\extrarowheight| (currently defined in
the \texttt{array} package) or \verb|\hhline| (currently defined in
the \texttt{hhline} package).  Note that you would not declare
\verb|\@tabularcr|, as this is an internal command, and the fact that
it is undefined in the \texttt{array.sty} version does not constitute
a change in the interface.

After that you may have any series of \LaTeX{} commands, these should
be called in such a way that any incorrect behaviour would show up in
the \texttt{.log} file.

Any commands, such as those allocating registers, which produce lines
in the \texttt{.log} file, which you do not want to be considered as part
of the test, you should surround with the declarations \verb|\OMIT|
and \verb|\TIMO|.

Finally the file should finish with either \verb|\END|, or
\verb|\end{document}|. Only use the second form if you actually need to
close the \texttt{document} environment, say to get the \texttt{.aux} files
correct.

\section{Tips}
The test should be written to show that the command meets its
specification. Ideally it should not fail if the command is correctly
re-implemented in a different way. This means that the primitive tracing
facilities like \verb|\tracingmacros| and \verb|\tracingcommands| should
probably not be used, as these reveal the implementation details of the
command.  

If the command involves visual formatting, one way to get information
into the \texttt{.log} file is to do all the commands inside
\verb|\setbox0=\vbox{...}|, and then \verb|\showbox0|. Before showing 
the box you should set \verb|\showboxdepth| and \verb|\showboxbreadth|
to suitable values so that just the right amount of information
displayed. \LaTeX{} has a command called \verb=\showoutput= which can
be of some help here.

\verb|\tracingoutput| or \verb|\tracingpages| might also be used (with
care), say if you were testing section headings at the top and bottom
of pages.  Other useful commands are \verb|\typeout{\meaning...}|,
which is like
\verb|\show|, but puts less junk in the \texttt{.log} file, and
\verb|\typeout{\the...}|, which is like \verb|\showthe|. 

Please remember that it isn't sufficient to produce a test file that
shows the desired behavior in the {\tt.dvi} file since this file will
not be inspected by the test system later on. Nevertheless, it is
often helpful to check the {\tt.dvi} when creating a test file but you
have to make sure that the relevant information is also displayed in
the {\tt.tlg} in the end.


\section{What format to use}

The format used to run the tests should not contain local specialties,
i.e., it should have been built from a standard distribution using the
no configuration files for fonts (i.e., no \texttt{fonttext.cfg} and
no \texttt{fontmath.cfg}) nor a configuration file for hyphenation
patterns (i.e., no \texttt{hyphen.cfg}). In particular, this means
that the format will have only standard American hyphenation patterns
included and that it will not support Babel extensions.

At some time in the future multi-lingual extensions should be added to
the scope of the test suite but we prefer to wait with that for the
release of a new multi-lingual interface to the kernel,
see~\cite{tub:MR97} for research results on that interface.


\section{Creating the .tlg File} 


If you are on a UNIX machine, a supplied shell script will run {\tt
  latex} twice on your test file, then edit the \texttt{.log} to
produce a \texttt{.tlg} file automatically.  At present, on other
operating systems, you will need to do this by hand.\footnote{We are
  grateful if we can include scripts for other operating systems to do
  this task.} Run \LaTeX{} twice, afterwards delete all lines upto and
including \verb|START-TEST-LOG|, and all lines after, and including
\verb|END-TEST-LOG|.  Similarly remove lines \verb|OMIT| and
\verb|TIMO|, and everything in between.  In addition remove all lines
that refer to loading a font definition file (extension \texttt{.fd})
including the line containing the closing parentheses denoting that
the file was closed.  Finally replace the base part of the test file
name whenever it shows up with ``nothing'', e.g., turn
`\texttt{(tlen01.aux)}' into `{\tt(.aux)}'. In this way the files stay
valid if we have to rename them.

\section{Sending new files to the \LaTeX{} maintainers}

It is important to realize, that only completely documented and
checked files can be included into the automated checking system for
\LaTeX2.09. The documentation, i.e., information what is supposed to
show up in the {.tlg} file, etc., is very important, since if later
such a file shows a change we need to be able to determine the reason
for it. It is also important that we don't have to re-check the test
file, in other words we like to get both the test file and the
hand-check result file, i.e., the {\tt.tlg} file from you. Please send
them to the address mentioned in \texttt{readme.val} file.


\section{If You Find a Bug}
The basic idea of this test system is to ensure that future releases
do not introduce new bugs. That is, it will normally be assumed that
if the new release produces different results on a test to an old
release, then the new release has a bug. Because of this, if you
discover while producing the file, that \LaTeX{} acts incorrectly in
some context, still submit the file, but note clearly in the file, and
in the email message, that you consider the current behaviour
incorrect. Otherwise you may ensure that this behaviour is maintained
in all future releases!


\section{Call for volunteers}

The more test files there are within this test suite exists the more
reliable the \LaTeX{} system will become. We therefore call on you for
help!  What we hope to find is a new group of volunteers that is
interested in working on an extension of the \LaTeX{} test suite
system. There is no need to be an expert \TeX{} or \LaTeX{} programmer
for this task though some experience with \LaTeX{} and its inner
workings will be necessary.

\begin{center}
\fbox{\begin{minipage}{\linewidth-2em}
  \bfseries If you are interested in joining this effort, please
  contact Daniel Flipo at 
\begin{center}
  \texttt{Daniel.Flipo@univ-lille1.fr}
\end{center}
  who kindly agreed to act as a coordinator between the individual
  volunteers.
\end{minipage}}
\end{center}
Further details about this effort can be found in~\cite{tub:xxx}.

\section{Sample files}

As an example we show a test file for the ``length'' module of
\LaTeX2.09.
Please note that it is important to document the desired output and
also that only relevant information shows up in the {\tt.tlg} file.
For this reason the allocation information for \verb=\newlength= is
suppressed since it may change without making the interface invalid.

\subsection{The file \texttt{tlen01.lvt}}
\verbfile{tlen01.lvt}

\onecolumn

\subsection{The file \texttt{tlen01.tlg}}
This is the resulting output after applying the UNIX script {\tt
maketlg} to the file \texttt{tlen01.lvt}.
\verbfile{tlen01.tlg}



\begin{thebibliography}{1}
\bibitem{A-W:LLa94}
Leslie Lamport.
\newblock {\em {\LaTeX:} A Document Preparation System}.
\newblock Addison-Wesley, Reading, Massachusetts, second edition, 1994.

\bibitem{tub:xxx}
Frank Mittelbach.
\newblock \emph{A regression test suite for \LaTeXe}.
\newblock \TUB{} (?)\ldots

\bibitem{tub:MR97}
Frank Mittelbach and Chris Rowley.
\newblock \emph{Language Information in   Structured Documents:
  A Model for Mark-up and Rendering}.
\newblock \TUB{} (?)\ldots

\end{thebibliography}



\end{document}


